---
title: "Assessment 2"
author: "Marlies Smit and Olivia Veal"
date: "2026-04-01"
output:
  bookdown::html_document2:
    number_sections: false
    

---

## Question 3:
How does each region compare in access to clean fuels for cooking?
The following section examines the difference in access to clean fuels for cooking for each region.


```{r Loading libraries and data, include=FALSE}
library(bookdown)
library(dplyr)
library(tidyr)
library(knitr)
library(tidyverse)
library(ggplot2)
library(plotly)
library(scales)

library(viridis)
# colours were selected from the viridis pallete which is a colour blind friendly pallete
colors = c("#E16462FF", "#6A00A8FF", "#FCA636FF", "#0D0887FF", "#B12A90FF",  "#FF60A8", "#C05780")

# Get rid of scientific notation
options(scipen = 999)

options(digits = 2)


csv1 = read.csv("Data/access-to-clean-fuels-for-cooking-vs-gdp-per-capita.csv")
# csv
```



```{r csv clean data}
csv2 <- csv1 %>%
  rename("Access" = "Access.to.clean.fuels.and.technologies.for.cooking") %>%
  rename("Country" = "Entity") %>%
  rename("GDP_per_capita" = "GDP.per.capita") %>%
  rename("OWID_world_region" = "World.region.according.to.OWID")
```

```{r Dealing with missing data}
# Getting 2023 data for each region
reg <- subset(csv2, Year == "2000" | Year == "2023")
reg <- subset(reg, OWID_world_region != "")

reg <- reg %>%
  # Select essential columns
  select(Country, Code, Year, Access, OWID_world_region)

# # returns data that does not contain (UN) or (WHO)
reg <- reg[!grepl("(UN)|(WHO)", reg$Country),]


# identify missing data
reg_filtered_out <- reg %>% 
  filter(if_any(c(Code, OWID_world_region, Access), ~is.na(.)))


# Data that is complete for analysis
# Only countries will have a Code and OWID_world_region
reg_filtered_in <- reg %>% drop_na()
# reset index
rownames(reg_filtered_in) <- NULL


# For reference countries that were filtered in or out
country_filtered_out <- unique(reg_filtered_out$Country)
# print(country_filtered_out)
country_filtered_in <- unique(reg_filtered_in$Country)
# print(country_filtered_in)

reg <- reg_filtered_in
```


```{r Formatting data for figure 3.1}
reg_wide <- reg %>%
  # pivot wider for the years
  pivot_wider(
    names_from = Year,
    values_from = Access
  )

# # create new Access change column
# # backticks to perform calculations for columns with number titles
reg_wide <- transform(reg_wide, Access_change = `2023` - `2000`)
```

<br>

### Question 3 table - Access statistics by region

The following table provides summary statistics to compare each region.
It shows the max, min and mean for each region in 2023, the change of the average in the past 15 years, and what percentage of the region was above the global average of `r avg_2023`%.

```{r Creating 3.1 table, message = FALSE}
# We want the mean
reg_mean <- reg_wide %>%
  group_by(OWID_world_region) %>%
  summarise(
    Max_2023 = max(`2023`),
    Min_2023 = min(`2023`),
    Mean_2023 = mean(`2023`),
    Mean_change = mean(Access_change)
  )
# reg_mean

# Calculate which percentage of countries are above the global average for 2023
# get the number of countries in each region
reg_count <- reg_wide %>%
  group_by(OWID_world_region) %>%
  summarise(
    Country_count = n_distinct(Country)
  )

# get global average for 2023
avg_2023 <- mean(reg_wide$`2023`)

# create a new column
avg_compare <- reg_wide %>%
  mutate(Compared_to_average = if_else(`2023` > avg_2023, "Above", "Below"))
# avg_compare

# Get the number of countries in each region that were above average
above_count <- avg_compare %>%
  group_by(OWID_world_region, Compared_to_average) %>%
  summarise(
    Above_count = n_distinct(Country)
  ) %>% filter(
    Compared_to_average == "Above"
  )
  

reg_merge1 <- merge(above_count, reg_count)
reg_merge2 <- merge(reg_merge1, reg_mean)

# get percentage above average
reg_merge3 <- transform(reg_merge2, Percentage_above_average = (Above_count/Country_count)*100)

# convert to percentage with two decimal places
reg_merge4 <- reg_merge3 %>%
  mutate(across(where(is.numeric), ~ .x / 100)) %>%
  mutate(across(where(is.numeric), label_percent(accuracy = 0.01)))
  

# select final columns for table
reg_table <- reg_merge4 %>%
  select(-c(Compared_to_average, Above_count, Country_count))
# reg_table

```


```{r, label = "table-access-stats"}
# table for regional access
kable(reg_table, col.names = c(
  "Region",
  "Max",
  "Min",
  "Mean",
  "Mean change since 2000",
  "Percentage above average"),
  format = "pipe", padding = 15,
  caption = "Access statistics by region for 2023",
  align = "r"
  )
```

```{r key statistics}

Max_Europe <- reg_table[3, "Mean_2023"]

Change_Asia <- reg_table[2, "Mean_change"]
```


<br>

### Question 3 figure - Regional average boxplot

The following box plot provided a visual representation of the range of access within each region. The colour of each point shows which quartile every country falls in country falls in regardless of region. By hovering over each country, the percentage of access and the country name is shown.

```{r 3.2 boxplot, fig.width=9, fig.height= 5, warning = FALSE, fig.align='center'}
reg_qtrs <- reg_wide %>%
  mutate(Quartiles = ntile(`2023`, 4)) %>%
  mutate(
    Quartiles = case_when(
    Quartiles == 1 ~ "First quartile",
    Quartiles == 2 ~ "Second quartile",
    Quartiles == 3 ~ "Third quartile",
    Quartiles == 4 ~ "Fourth quartile")) %>%
  mutate(
    Quartiles_ordered = factor(Quartiles,
               levels = c(
                 "First quartile",
                 "Second quartile",
                 "Third quartile",
                 "Fourth quartile"))
  )

reg_box <- reg_qtrs  %>%
  ggplot(aes(y = `2023`, x = OWID_world_region, label = Country)) +
  geom_boxplot() +
  geom_jitter(aes(colour = Quartiles_ordered)) +
  theme_light() +
  coord_flip() +
  scale_color_manual(values = colors) +
  labs(x = NULL, y = "Access (%)",
       colour = "Global quartiles") +
  scale_y_continuous(breaks = seq(0, 100, by = 10))


fig_3.2 <- ggplotly(reg_box, tooltip = c("Country", "2023"))
```

```{r reg-ave-plot, fig.cap= "Regional average boxplot",  fig.width=9, fig.height= 5, warning = FALSE, fig.align='center'}
fig_3.2
```


 *The black dots with no country represent the quartiles*
 

## Question 4:
Are there countries that have high access to clean fuels relative to their GDP per capita, and do they have anything in common?

### Question 4 table - Top ten countries with high access compared to their GDP per capita

```{r identifying outliers with residuals}
# create a linear regression model

# clean data
relation1 <- merge(reg, csv2)
relation2 <- subset(relation1, Year == "2023")
# 10 countries are missing GDP per capita data
relation3 <- relation2[!is.na(relation2$GDP_per_capita), ]

# reset index
rownames(relation2) <- NULL

# create a regression model
model <- lm(Access ~ GDP_per_capita, data = relation3)

# residuals tell how far each point is from the regression line
relation3$Residuals <- resid(model)

relation <- relation3
# relation
```


```{r Get the top ten}
high_resid1 <- relation %>% 
  select(Country, Access, GDP_per_capita, Residuals, OWID_world_region, Population) %>%
  # arrange by Residuals
  arrange(desc(Residuals))


high_resid2 <- high_resid1 %>%
  mutate(across(Access, ~ .x / 100)) %>%
  mutate(across(Access, label_percent(accuracy = 0.01))) %>%
  mutate(GDP_per_capita = round(GDP_per_capita, 2)) %>%
  mutate(GDP_per_capita = paste0("$", GDP_per_capita))

# reg_merge4 <- reg_merge3 %>%
#   mutate(across(where(is.numeric), ~ .x / 100)) %>%
#   mutate(across(where(is.numeric), label_percent(accuracy = 0.01)))

high_resid <- high_resid2
# high_resid

top_ten <- head(high_resid, 10)
```


```{r label = "top-ten-table"}

# table for regional access
kable(top_ten, col.names = c(
  "Country",
  "Access",
  "GDP per capita",
  "Residuals",
  "Region",
  "Population"),
  format = "pipe", padding = 15,
  caption = "Top ten countries with high access for their GDP"
  )

```


<br>

### Question 4 figure - Distribution of relative high access for GDP per capita

The following visualisation further investigated countries with a GDP per capita under $17,000 similar to the top ten countries from Figure 4.1. The scatterplot showed the region, GDP per capita, and residuals from our regression analysis for each country and indicated the population by the size of each point.


```{r Create grpahic 4.2}
# get only countries with a similar GDP
resid_vis1 <- subset(high_resid1, GDP_per_capita <= 17000)

# change the index to Country for the label
rownames(resid_vis1) <- resid_vis1$Country

# Put access into percentage format
resid_vis2 <- resid_vis1 %>%
  mutate(across(Access, ~ .x / 100)) %>%
  mutate(across(Access, label_percent(accuracy = 0.01)))

# Create custom label for interactive plot to see country and access (since they aren't variables in the resid_distro ggplot graph)
resid_vis <- resid_vis2 %>%
  mutate(country_access = paste0("Country: ", rownames(resid_vis1),
                                 "<br>Access: ", Access,
                                 "<br>Population: ", Population))

resid_distro <- ggplot(resid_vis, aes(y = Residuals,
                      x = GDP_per_capita,
                      size = Population,
                      fill = OWID_world_region,
                      text = country_access)) +
  geom_point(alpha = 0.5, shape=21, color= "black")+
  labs(x = "GDP per capita ($)", y = "Residuals", fill = "Region") +
  scale_fill_manual(values = colors) +
  scale_size(range = c(1, 20), name = " ")+
  theme_light()


resid_final <- ggplotly(resid_distro, tooltip = "text")
```
```{r resid-plot, fig.cap= "Residuals distribution",fig.width=9, fig.height= 5, warning = FALSE, fig.align='center'}
resid_final
```


<br>

<br>
  
## Section analysis for Questions three and four:


In this section we investigated how region shapes access, and tried to develop an understanding of what characteristics if any countries that had relatively high access for their wealth (measured in GDP per capita) had in common. The purpose was to identify factors that possible contribute to a high percentage of access for the population, and identify countries with conditions that are appropriate for investment into accessible clean fuels for cooking.

From our analysis we found that different regions had vastly different distribution of access between countries. The level of access within a region was mapped to the level of wealth. Regions that were wealthier had more countries with a level of access at 100%. As seen in Figure \@ref(fig:resid-plot), all regions had a average level of access above the global average (`r avg_2023`%) for 2023, except Africa. Africa only had 18.87% of countries above the global average seen in Table \@ref(tab:table-access-stats).

Following on from the regression analysis to determine the relationship between GDP per capita and Access we drilled down on the countries that had the highest level of access for their wealth measured in the positive distance from the regression line or positive residual. Table \@ref(tab:top-ten-table) showed that the top ten countries had a GDP per capita between $8000 and \$17000. Figure \@ref(fig:resid-plot) showed that countries with positive residuals did not have a shared region or population size.