---
title: "Assessment 2"
author: "Marlies Smit and Olivia Veal"
date: "2026-04-01"
output:
  bookdown::html_document2:
    number_sections: false

---

## Question 3:
How does each region compare in access to clean fuels for cooking?
The following section examines the difference in access to clean fuels for cooking for each region.


```{r Loading libraries and data, include=FALSE}
library(bookdown)
library(dplyr)
library(tidyr)
library(knitr)
library(tidyverse)
library(ggplot2)
library(plotly)
library(scales)

library(viridis)
# colours were selected from the viridis pallete which is a colour blind friendly pallete
colors = c("#E16462FF", "#6A00A8FF", "#FCA636FF", "#0D0887FF", "#B12A90FF",  "#FF60A8", "#C05780")

# Get rid of scientific notation
options(scipen = 999)

options(digits = 3)


csv = read.csv("Data/access-to-clean-fuels-for-cooking-vs-gdp-per-capita.csv")
# csv
```

<br>

### Figure 3.1 - Access statistics by region


```{r csv viewing}
csv <- csv %>%
  rename("Access" = "Access.to.clean.fuels.and.technologies.for.cooking") %>%
  rename("GDP_per_capita" = "GDP.per.capita") %>%
  rename("OWID_world_region" = "World.region.according.to.OWID")
```

```{r Dealing with missing data}
# Getting 2023 data for each region
reg <- subset(csv, Year == "2000" | Year == "2023")
reg <- subset(reg, OWID_world_region != "")

reg <- reg %>%
  # Select essential columns
  select(Entity, Code, Year, Access, OWID_world_region)

# # returns data that does not contain (UN) or (WHO)
reg <- reg[!grepl("(UN)|(WHO)", reg$Entity),]


# identify missing data
reg_filtered_out <- reg %>% 
  filter(if_any(c(Code, OWID_world_region, Access), ~is.na(.)))


# Data that is complete for analysis
# Only countries will have a Code and OWID_world_region
reg_filtered_in <- reg %>% drop_na()
# reset index
rownames(reg_filtered_in) <- NULL


# For reference countries that were filtered in or out
country_filtered_out <- unique(reg_filtered_out$Entity)
# print(country_filtered_out)
country_filtered_in <- unique(reg_filtered_in$Entity)
# print(country_filtered_in)

reg <- reg_filtered_in
```


```{r Formatting data for figure 3.1}
reg_wide <- reg %>%
  # pivot wider for the years
  pivot_wider(
    names_from = Year,
    values_from = Access
  )

# # create new Access change column
# # backticks to perform calculations for columns with number titles
reg_wide <- transform(reg_wide, Access_change = `2023` - `2000`)

# reg_wide
```

```{r Creating 3.1 table, message = FALSE}
# We want the mean
reg_mean <- reg_wide %>%
  group_by(OWID_world_region) %>%
  summarise(
    Max_2023 = max(`2023`),
    Min_2023 = min(`2023`),
    Mean_2023 = mean(`2023`),
    Mean_change = mean(Access_change)
  )
# reg_mean

# Calculate which percentage of countries are above the global average for 2023
# get the number of countries in each region
reg_count <- reg_wide %>%
  group_by(OWID_world_region) %>%
  summarise(
    Country_count = n_distinct(Entity)
  )

# get global average for 2023
avg_2023 <- mean(reg_wide$`2023`)

# create a new column
avg_compare <- reg_wide %>%
  mutate(Compared_to_average = if_else(`2023` > avg_2023, "Above", "Below"))
# avg_compare

# Get the number of countries in each region that were above average
above_count <- avg_compare %>%
  group_by(OWID_world_region, Compared_to_average) %>%
  summarise(
    Above_count = n_distinct(Entity)
  ) %>% filter(
    Compared_to_average == "Above"
  )
  

reg_merge1 <- merge(above_count, reg_count)
reg_merge2 <- merge(reg_merge1, reg_mean)

# get percentage above average
reg_merge3 <- transform(reg_merge2, Percentage_above_average = (Above_count/Country_count)*100)

# convert to percentage with two decimal places
reg_merge4 <- reg_merge3 %>%
  mutate(across(where(is.numeric), ~ .x / 100)) %>%
  mutate(across(where(is.numeric), label_percent(accuracy = 0.01)))
  

# select final columns for table
reg_table <- reg_merge4 %>%
  select(-c(Compared_to_average, Above_count, Country_count))
# reg_table

```

The following table provides summary statistics to compare each region.
It shows the max, min and mean for each region in 2023, the change of the average in the past 15 years, and what percentage of the region was above the global average of `r avg_2023`%.


```{r Figure 3.1 table}
# table for regional access
kable(reg_table, col.names = c(
  "Region",
  "Max",
  "Min",
  "Mean",
  "Mean change since 2000",
  "Percentage above average"),
  format = "pipe", padding = 15,
  caption = "Figure 3.1 - Access statistics by region for 2023",
  align = "r"
  )
```

```{r key statistics}

Max_Europe <- reg_table[3, "Mean_2023"]

Change_Asia <- reg_table[2, "Mean_change"]
```


From the table we notice the following:

- All regions had at least one country with 100% access
- Europe had the highest level of access on average at `r Max_Europe`
- South America was the only region with all countries above the 2023 global average (`r avg_2023`%)
- Asia had the greatest increase to access at `r Change_Asia` while all other regions were less than 20%
- Africa has the lowest Min, Mean and Percentage above average



<br>

### Figure 3.2 - Regional average boxplot

The following box plot will provide a visual representation of the range of access within each region. The colour of each point shows which quartile every country falls in country falls in regardless of region. By hovering over each country, the percentage of access and the country name is shown.

```{r 3.2 boxplot, fig.width=9, fig.height= 5, warning = FALSE, fig.align='center'}
reg_qtrs <- reg_wide %>%
  mutate(Quartiles = ntile(`2023`, 4)) %>%
  mutate(
    Quartiles = case_when(
    Quartiles == 1 ~ "First quartile",
    Quartiles == 2 ~ "Second quartile",
    Quartiles == 3 ~ "Third quartile",
    Quartiles == 4 ~ "Fourth quartile")) %>%
  mutate(
    Quartiles_ordered = factor(Quartiles,
               levels = c(
                 "First quartile",
                 "Second quartile",
                 "Third quartile",
                 "Fourth quartile"))
  )

reg_box <- reg_qtrs  %>%
  ggplot(aes(y = `2023`, x = OWID_world_region, label = Entity)) +
  geom_boxplot() +
  geom_jitter(aes(colour = Quartiles_ordered)) +
  theme_light() +
  coord_flip() +
  scale_color_manual(values = colors) +
  labs(x = NULL, y = "Access (%)",
       title = "Figure 3.2 - Regional average boxplot",
       colour = "Global quartiles") +
  scale_y_continuous(breaks = seq(0, 100, by = 10))


fig_3.2 <- ggplotly(reg_box, tooltip = c("Entity", "2023"))
fig_3.2
```

 *The black dots with no country represent the quartiles*
 
 Figure 3.2 provided the following insights:
 
 - To be in the highest quartile countries need to have 100% access
 - Europe had the smallest interquartile range, whereas Oceania had the largest
 - Africa was the only region to have their average (as well a their third quartile) below the global average
 

## Question 4:
Are there countries that have high access to clean fuels relative to their GDP per capita, and do they have anything in common?

How do you determine if something is high compared to another variable?
