---
title: "Assessment 2"
author: "Marlies Smit and Olivia Veal"
date: "2026-04-01"
output:
  bookdown::html_document2:
    number_sections: false
bibliography: references.bib
link-citations: true
---

### Introduction

GDP per capita plays an integral role in shaping the capacity of households and governments to invest in access to clean cooking technologies. Research on access has showed that GDP per capita plays an integral role in shaping the capacity of households and governments to invest in access to clean cooking technologies. @onyeneke2023 showed that a 1% increase in GDP per capita significantly increased access to clean fuels by 0.219% in Africa alone. As well as showing a positive correlation between GDP growth and clean energy access. This study underscores that populations with higher disposable income are generally better positioned to have access to cleaner alternatives in fuels [@scott2023]. However, there are additional complementary factors that needs to be considered, one of which may be population. A study from @malahkuete2025 shows that some higher population densities create favourable conditions for access to clean fuels, such as regions in Southeast Asia. Thereby highlighting the need for investigation of multiple interacting factors with population and GDP.

This report aimed to build on research regarding how different demographic factors influence a country’s level of access to clean fuels for cooking. The report analysed data provided by @owid (processed by Our World of Data) to examine the correlation between access to clean cooking fuels and the factors: GDP per capita, population size and region. Statistical methods such as linear regression analysis were employed to determine the significance of the relationship such as GDP per capita and access, as well as assess if a country had a high level of access for their GDP per capita in relation to other countries. To investigate the current academic research, the following four questions were posted: - What is the relationship between GDP per capita and access to clean fuels? - What is the relationship between population size and access to clean fuels? - How does each region compare in access to clean fuels for cooking? - Are there countries that have high access to clean fuels relative to their GDP per capita, and do they have anything in common?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(tidyr)
library(dplyr)
library(scales)
library(ggplot2)
library(knitr)

```

### Question 1: What is the relationship between GDP per capita and access to clean fuels?

```{r}
access_data<-read.csv("data/access-to-clean-fuels-for-cooking-vs-gdp-per-capita.csv")
```

```{r}
kable(head(access_data), caption = "Variables of the Dataset")
```

```{r}
clean_data <- access_data %>%
  select(Entity, Year, `GDP.per.capita`, Population, `Access.to.clean.fuels.and.technologies.for.cooking`) %>%
  drop_na()
```

#### Question 1 Figure - Relationship Scatterplot

```{r gg-GDP-plot, fig.cap = "Scatterplot investigating the relationship between GDP per capita and access to clean fuels"}
ggplot(clean_data, aes(x = GDP.per.capita, y = Access.to.clean.fuels.and.technologies.for.cooking)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  scale_x_log10() +
  labs(title = "Relationship between GDP per capita and Access to Clean Fuels",
       x = "GDP per capita (log scale)",
       y = "Access to Clean Fuels (%)") +
  theme_light()
```

```{r}
# Correlation
cor(clean_data$`GDP.per.capita`, clean_data$`Access.to.clean.fuels.and.technologies.for.cooking`, use = "complete.obs")
```

#### Question 1 Table - Correlation Statistics

```{r tab:sum-table}
summary_stats <- data.frame(
  Mean_GDP = mean(clean_data$GDP.per.capita),
  SD_GDP = sd(clean_data$GDP.per.capita),
  Mean_Access = mean(clean_data$Access.to.clean.fuels.and.technologies.for.cooking),
  SD_Access = sd(clean_data$Access.to.clean.fuels.and.technologies.for.cooking),
  Correlation = cor(clean_data$GDP.per.capita, clean_data$Access.to.clean.fuels.and.technologies.for.cooking)
)

kable(summary_stats, caption = "Summary Statistics and Correlation",
      label = "sum-table")
```

The scatterplot and correlation investigating the relationship between GDP per capita and access to clean fuels shows 0.63, a moderately strong positive relationship. What this shows is as GDP per capita increases, the access to clean fuels also tends to increase (See Figure \@ref(fig:gg-GDP-plot)).This correlation can also be seen in the correlation table between the two variables (See Table \@ref(tab:sum-table)).

The value of 0.63 being moderately strong is a good, fairly consistent predictor, however, not perfect. What this may show is that some countries will have high GDP but low access or vice versa. However, generally, wealthier countries will have better infrastructure and energy access allowing for more clean fuels. Other explanatory factors will most likely be able to have stronger, if not additional, predictors for the relationship.

### Question 2: What is the relationship between population size and access to clean fuels?

#### Question 2 Figure - Relationship Scatterplots

```{r gg-Pop-plot, fig.cap = "Scatterplot investigating the relationship between population and access to clean fuels"}
ggplot(clean_data, aes(x = Population, y = Access.to.clean.fuels.and.technologies.for.cooking)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  scale_x_log10() +
    labs(title = "Relationship between Population and Access to Clean Fuels",
       x = "Population (log scale)",
       y = "Access to Clean Fuels (%)") +
  theme_light()
```

```{r gg-Hex-plot, fig.cap = "Scatterplot investigating the relationship between population and access to clean fuels in a hex map"}
ggplot(clean_data, aes(x = Population, y = `Access.to.clean.fuels.and.technologies.for.cooking`)) +
  geom_hex(bins = 30) +
  scale_x_log10() +
  scale_fill_continuous(type = "viridis") +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  labs(title = "Population vs Access to Clean Fuels (Density Plot)",
       x = "Population (log scale)",
       y = "Access to Clean Fuels (%)") +
  theme_minimal()
```

```{r}
# Correlation
cor(clean_data$`Population`, clean_data$`Access.to.clean.fuels.and.technologies.for.cooking`, use = "complete.obs")
```

#### Question 2 Table - Correlation Statistics

```{r tab:sum-table1}
summary_stats1 <- data.frame(
  Mean_Population = mean(clean_data$Population),
  SD_Population = sd(clean_data$Population),
  Mean_Access = mean(clean_data$Access.to.clean.fuels.and.technologies.for.cooking),
  SD_Access = sd(clean_data$Access.to.clean.fuels.and.technologies.for.cooking),
  Correlation = cor(clean_data$Population, clean_data$Access.to.clean.fuels.and.technologies.for.cooking)
)

kable(summary_stats1, caption = "Summary Statistics and Correlation",
      label = "sum-table1")
```

Investigating the relationship between population and access to clean fuels comparatively is quite weak and this is shown by the weak negative relationship of -0.01. This correlation can also be seen in the correlation table between the two variables (See Figure \@ref(tab:sum-table1)). From the scatterplot it can be seen that the correlation line slopes slightly downward, echoing the weak negative relationship, indicating that countries with larger population tends to have slightly lower access to clean fuels on average, but the effect is not strong (See Figure \@ref(fig:gg-Pop-plot)).

For this relationship investigation the scatter plot shows a number of vertical stripes. This is shown as there are many countries with similar population sizes and differing accesses to clean fuels. A hex plot was also selected to indicate how many countries share similar population sizes in areas of access to clean fuels (See Figure \@ref(fig:gg-Hex-plot)). This relationship shown is a clear indication that population alone cannot explain much of the variation in access to clean fuels, much like the academic research indicates.

### Question 3: How does each region compare in access to clean fuels for cooking?

```{r Loading libraries and data, include=FALSE}
library(bookdown)
library(dplyr)
library(tidyr)
library(knitr)
library(tidyverse)
library(ggplot2)
library(plotly)
library(scales)

library(viridis)
# colours were selected from the viridis pallete which is a colour blind friendly pallete
colors = c("#E16462FF", "#6A00A8FF", "#FCA636FF", "#0D0887FF", "#B12A90FF",  "#FF60A8", "#C05780")

# Get rid of scientific notation
options(scipen = 999)

options(digits = 2)


csv1 = read.csv("Data/access-to-clean-fuels-for-cooking-vs-gdp-per-capita.csv")
# csv
```

```{r csv clean data}
csv2 <- csv1 %>%
  rename("Access" = "Access.to.clean.fuels.and.technologies.for.cooking") %>%
  rename("Country" = "Entity") %>%
  rename("GDP_per_capita" = "GDP.per.capita") %>%
  rename("OWID_world_region" = "World.region.according.to.OWID")
```

```{r Dealing with missing data}
# Getting 2023 data for each region
reg <- subset(csv2, Year == "2000" | Year == "2023")
reg <- subset(reg, OWID_world_region != "")

reg <- reg %>%
  # Select essential columns
  select(Country, Code, Year, Access, OWID_world_region)

# # returns data that does not contain (UN) or (WHO)
reg <- reg[!grepl("(UN)|(WHO)", reg$Country),]


# identify missing data
reg_filtered_out <- reg %>% 
  filter(if_any(c(Code, OWID_world_region, Access), ~is.na(.)))


# Data that is complete for analysis
# Only countries will have a Code and OWID_world_region
reg_filtered_in <- reg %>% drop_na()
# reset index
rownames(reg_filtered_in) <- NULL


# For reference countries that were filtered in or out
country_filtered_out <- unique(reg_filtered_out$Country)
# print(country_filtered_out)
country_filtered_in <- unique(reg_filtered_in$Country)
# print(country_filtered_in)

reg <- reg_filtered_in
```

```{r Formatting data for figure 3.1}
reg_wide <- reg %>%
  # pivot wider for the years
  pivot_wider(
    names_from = Year,
    values_from = Access
  )

# # create new Access change column
# # backticks to perform calculations for columns with number titles
reg_wide <- transform(reg_wide, Access_change = `2023` - `2000`)
```

#### Question 3 Table - Access statistics by region

```{r Creating 3.1 table, message = FALSE}
# We want the mean
reg_mean <- reg_wide %>%
  group_by(OWID_world_region) %>%
  summarise(
    Max_2023 = max(`2023`),
    Min_2023 = min(`2023`),
    Mean_2023 = mean(`2023`),
    Mean_change = mean(Access_change)
  )
# reg_mean

# Calculate which percentage of countries are above the global average for 2023
# get the number of countries in each region
reg_count <- reg_wide %>%
  group_by(OWID_world_region) %>%
  summarise(
    Country_count = n_distinct(Country)
  )

# get global average for 2023
avg_2023 <- mean(reg_wide$`2023`)

# create a new column
avg_compare <- reg_wide %>%
  mutate(Compared_to_average = if_else(`2023` > avg_2023, "Above", "Below"))
# avg_compare

# Get the number of countries in each region that were above average
above_count <- avg_compare %>%
  group_by(OWID_world_region, Compared_to_average) %>%
  summarise(
    Above_count = n_distinct(Country)
  ) %>% filter(
    Compared_to_average == "Above"
  )
  

reg_merge1 <- merge(above_count, reg_count)
reg_merge2 <- merge(reg_merge1, reg_mean)

# get percentage above average
reg_merge3 <- transform(reg_merge2, Percentage_above_average = (Above_count/Country_count)*100)

# convert to percentage with two decimal places
reg_merge4 <- reg_merge3 %>%
  mutate(across(where(is.numeric), ~ .x / 100)) %>%
  mutate(across(where(is.numeric), label_percent(accuracy = 0.01)))
  

# select final columns for table
reg_table <- reg_merge4 %>%
  select(-c(Compared_to_average, Above_count, Country_count))
# reg_table

```

The following table provides summary statistics to compare each region. It shows the max, min and mean for each region in 2023, the change of the average in the past 15 years, and what percentage of the region was above the global average of `r avg_2023`%.

```{r, label = "table-access-stats"}
# table for regional access
kable(reg_table, col.names = c(
  "Region",
  "Max",
  "Min",
  "Mean",
  "Mean change since 2000",
  "Percentage above average"),
  format = "pipe", padding = 15,
  caption = "Access statistics by region for 2023",
  align = "r"
  )
```

```{r key statistics}

Max_Europe <- reg_table[3, "Mean_2023"]

Change_Asia <- reg_table[2, "Mean_change"]
```

#### Question 3 Figure - Regional average boxplot

The following box plot provided a visual representation of the range of access within each region. The colour of each point shows which quartile every country falls in country falls in regardless of region. By hovering over each country, the percentage of access and the country name is shown.

```{r 3.2 boxplot, fig.width=9, fig.height= 5, warning = FALSE, fig.align='center'}
reg_qtrs <- reg_wide %>%
  mutate(Quartiles = ntile(`2023`, 4)) %>%
  mutate(
    Quartiles = case_when(
    Quartiles == 1 ~ "First quartile",
    Quartiles == 2 ~ "Second quartile",
    Quartiles == 3 ~ "Third quartile",
    Quartiles == 4 ~ "Fourth quartile")) %>%
  mutate(
    Quartiles_ordered = factor(Quartiles,
               levels = c(
                 "First quartile",
                 "Second quartile",
                 "Third quartile",
                 "Fourth quartile"))
  )

reg_box <- reg_qtrs  %>%
  ggplot(aes(y = `2023`, x = OWID_world_region, label = Country)) +
  geom_boxplot() +
  geom_jitter(aes(colour = Quartiles_ordered)) +
  theme_light() +
  coord_flip() +
  scale_color_manual(values = colors) +
  labs(x = NULL, y = "Access (%)",
       colour = "Global quartiles") +
  scale_y_continuous(breaks = seq(0, 100, by = 10))


fig_3.2 <- ggplotly(reg_box, tooltip = c("Country", "2023"))
```

```{r reg-ave-plot, fig.cap= "Regional average boxplot",  fig.width=9, fig.height= 5, warning = FALSE, fig.align='center'}
fig_3.2
```

*The black dots with no country represent the quartiles*

From the analysis it was found that different regions had vastly different distribution of access between countries. The level of access within a region was mapped to the level of wealth. Regions that were wealthier had more countries with a level of access at 100%. As seen in Figure \@ref(fig:reg-ave-plot), all regions had a average level of access above the global average (`r avg_2023`%) for 2023, except Africa. Africa only had 18.87% of countries above the global average seen in Table \@ref(tab:table-access-stats).

### Question 4: Are there countries that have high access to clean fuels relative to their GDP per capita, and do they have anything in common?

#### Question 4 Table - Top ten countries with high access compared to their GDP per capita

```{r identifying outliers with residuals}
# create a linear regression model

# clean data
relation1 <- merge(reg, csv2)
relation2 <- subset(relation1, Year == "2023")
# 10 countries are missing GDP per capita data
relation3 <- relation2[!is.na(relation2$GDP_per_capita), ]

# reset index
rownames(relation2) <- NULL

# create a regression model
model <- lm(Access ~ GDP_per_capita, data = relation3)

# residuals tell how far each point is from the regression line
relation3$Residuals <- resid(model)

relation <- relation3
# relation
```

```{r Get the top ten}
high_resid1 <- relation %>% 
  select(Country, Access, GDP_per_capita, Residuals, OWID_world_region, Population) %>%
  # arrange by Residuals
  arrange(desc(Residuals))


high_resid2 <- high_resid1 %>%
  mutate(across(Access, ~ .x / 100)) %>%
  mutate(across(Access, label_percent(accuracy = 0.01))) %>%
  mutate(GDP_per_capita = round(GDP_per_capita, 2)) %>%
  mutate(GDP_per_capita = paste0("$", GDP_per_capita))

# reg_merge4 <- reg_merge3 %>%
#   mutate(across(where(is.numeric), ~ .x / 100)) %>%
#   mutate(across(where(is.numeric), label_percent(accuracy = 0.01)))

high_resid <- high_resid2
# high_resid

top_ten <- head(high_resid, 10)
```

```{r label = "top-ten-table"}

# table for regional access
kable(top_ten, col.names = c(
  "Country",
  "Access",
  "GDP per capita",
  "Residuals",
  "Region",
  "Population"),
  format = "pipe", padding = 15,
  caption = "Top ten countries with high access for their GDP"
  )

```

#### Question 4 Figure - Distribution of relative high access for GDP per capita

The following visualisation further investigated countries with a GDP per capita under \$17,000 similar to the top ten countries from Figure 4.1. The scatterplot showed the region, GDP per capita, and residuals from our regression analysis for each country and indicated the population by the size of each point.

```{r Create grpahic 4.2}
# get only countries with a similar GDP
resid_vis1 <- subset(high_resid1, GDP_per_capita <= 17000)

# change the index to Country for the label
rownames(resid_vis1) <- resid_vis1$Country

# Put access into percentage format
resid_vis2 <- resid_vis1 %>%
  mutate(across(Access, ~ .x / 100)) %>%
  mutate(across(Access, label_percent(accuracy = 0.01)))

# Create custom label for interactive plot to see country and access (since they aren't variables in the resid_distro ggplot graph)
resid_vis <- resid_vis2 %>%
  mutate(country_access = paste0("Country: ", rownames(resid_vis1),
                                 "<br>Access: ", Access,
                                 "<br>Population: ", Population))

resid_distro <- ggplot(resid_vis, aes(y = Residuals,
                      x = GDP_per_capita,
                      size = Population,
                      fill = OWID_world_region,
                      text = country_access)) +
  geom_point(alpha = 0.5, shape=21, color= "black")+
  labs(x = "GDP per capita ($)", y = "Residuals", fill = "Region") +
  scale_fill_manual(values = colors) +
  scale_size(range = c(1, 20), name = " ")+
  theme_light()


resid_final <- ggplotly(resid_distro, tooltip = "text")
```

```{r resid-plot, fig.cap= "Residuals distribution",fig.width=9, fig.height= 5, warning = FALSE, fig.align='center'}
resid_final
```

Following on from the regression analysis to determine the relationship between GDP per capita and Access we drilled down on the countries that had the highest level of access for their wealth measured in the positive distance from the regression line or positive residual. Table \@ref(tab:top-ten-table) showed that the top ten countries had a GDP per capita between \$8000 and \$17000. Figure \@ref(fig:resid-plot) showed that countries with positive residuals did not have a shared region or population size.

In this section it was investigated how region shapes access, with a focus on developing an understanding of what characteristics if any countries that had relatively high access for their wealth (measured in GDP per capita) had in common. The purpose was to identify factors that possible contribute to a high percentage of access for the population, and identify countries with conditions that are appropriate for investment into accessible clean fuels for cooking.

### Conclusion:

This report had the aims of investigating various factors on access to clean fuels for cooking, by looking at data provided by the World Health Organisation for over 90 countries. Linear regression analysis on the significance of GDP per capita revealed that as GDP per capita increased access increased, yet the strength of the relationship was only 0.63 as seen in Table \@ref(tab:sum-table). Regional analysis revealed that this relationship may be weaker than expected because one quarter of the countries had 100% access (Figure \@ref(fig:reg-ave-plot)), and so as countries continued to get richer their access did not increase beyond a certain point.

Analysis was conducted on the correlation between population size and access which had a slight negative correlation as seen in Figure \@ref(fig:gg-Pop-plot). Analysis showed that population alone could not be an indicator of access to clean fuels for cooking and that other demographic factors may have influenced the findings the research by @malahkuete2025. In our analysis we saw that there was no commonality between countries with especially high access for their GDP per capita, which is outlined in Table \@ref(tab:top-ten-table).

This report echoed previous research finding that GDP per capita was the strongest indicator on the level of access in each country where population had minimal affect on access. It could be seen that regions were distinct, but no single region outperformed in having a high level of access for their countries' GDP per capita. With regard to predictions investiture in clean fuels, it may be seen that countries start to invest more in clean fuels for cooking when they reach their GDP per capita reaches the range between \$8000 and \$17000.

### References

\printbibliography
